---
title: "A Multilevel Bayesian Approach for Inferring the Effect of Fouls on Offensive Patterns in NBA Basketball Games"
execute:
  echo: false
  warning: false 
format:
  html:
    toc: false
    number-sections: true
    embed-resources: true
editor: visual
author: "Michael Light"
---

# Timeline

Editable [here](https://docs.google.com/spreadsheets/d/1-yxScU8IQDm0yY-b2s7P6D_-Z3uH5F8GrPLc_FQrKNs/edit?gid=0#gid=0).

```{r}
#| warning: false
#| output: true
library(googlesheets4)
library(gt)

timeline <- read_sheet("https://docs.google.com/spreadsheets/d/1-yxScU8IQDm0yY-b2s7P6D_-Z3uH5F8GrPLc_FQrKNs/edit?gid=0#gid=0")

timeline |> gt() |> 
    fmt_date(columns = date, date_style = "wd_m_day_year")

```

# Project Motivation

## Background on the Role of Physicality in NBA Games

Basketball is fundamentally a simple game: put a ball through a hoop. It's also a numbers game: put a ball through a hoop more times than your opponent. As a result, basketball analysis has historically focused on offensive aspects - who scored the most points, how they scored them, who assisted them, and so on. This offensive focus is natural given that the final score directly reflects which team scored more points.

The introduction of player tracking technology in the early 2010s briefly revolutionized defensive analysis by providing detailed data on defender positioning and movement. However, since 2016, this valuable player tracking data has been made private, creating a significant challenge for researchers outside NBA organizations.

This data limitation has created a critical gap in basketball analytics. Without access to detailed player tracking data, researchers are left with the same fundamental problem they faced before the tracking era: how can we effectively study and evaluate individual defensive ability when publicly available, "counting" statistics are either:

-   Primarily focused on offense

-   Only loosely related to defense (e.g., blocks and steals)

Solving this problem helps democratize defensive evaluation by creating metrics that can be derived from publicly available, play-by-play data. This would allow analysts outside NBA organizations to contribute meaningful insights to the game and help properly communicate the value of defensive players who may not show up in traditional box scores. Further, reliance on player tacking rather than play-by-play data relegates sample selection to a third-party data provider like [Synergy](https://sportradar.com/synergy-player-tracking/), whose publication aims may introduce bias by selecting observations which may not be representative of the game as a whole. As such, determining the viability of using counting stats to infer complex characteristics of games has the potential to reduce sytematic uncertainty and risk of future policy changes by third parties.

At a broader level, developing robust defensive metrics from public data could help bridge the growing divide between basketball analytics and fan understanding. Currently, there exists a significant gap between analysts who rely on advanced metrics and fans who primarily use traditional counting stats like points, rebounds, and assists. Developing novel, simple, and intuitive ways of understanding the impact of difficult-to-measure latent player characteristics like defensive skill represents a crucial challenge in modern basketball analysis.

## Purpose

The purpose of this research is to estimate the causal influence of individual defender foul counts on the shotmaking probability of opposing offenses under the hypothesis that players' defensive skill is a latent trait that can be estimated using NBA play-by-play data.

This work draws widely from fields including sports analytics, causal inference, Bayesian statistics, and item-response theory.

## Confounding Expected

Among the challenges quantifying any aspect of player skill—offensive, defensive, or otherwise—is separating individual from team contribution. For all the simplicity noted in the introduction, basketball isn't a game you play on paper (Simmons, 2009). We expect any analysis or model to be imprecise to some degree, and in this research attempt to account for unobserved confounds associated with shotmaking probability. These include offensive characteristics including offensive team, shot location, and shot type.

We narrow our study of defensive skill to offensive posessions that result in a shot. This means that we define defensive skill primarily as a component of any defending player's ability to affect the probability that any opposing offensive player successfully makes an attempted shot. We do not consider cases where an offensive possession results in a turnover, such as those caused by steals or shot clock violations, which although may be indicitative of good or bad defense being played are ignored for the purpose of this analysis.

As such, any estimate of defensive skill should be interpreted as an estimate of defensive skill in the context of shotmaking, and should be assumed independent of higher-level, latent traits like general defensive skill, general skill, offensive skill, etc.

## Missing Data

The dataset used for this paper include complete play-by-play records for every game played during the 2024-2025 NBA season. As our analysis relies on player-level position data reported on official NBA team rosters as of 2025-04-14, players who were included in games earlier in the season but were no longer part of any NBA rosters lacked position data and were dropped from the sample. This included TK players, accounting for TK percent of the raw data.

## Measurement

The target of inference (defensive skill) cannot be directly measured, and the primary measurement (foul count) is measured with some amount of error resulting from erroneous transcription in raw play_by_play data. Steps taken to account for transcription error can be found in the supplemental code.

Since foul counts differ between individual defender and defenders tend to play for fewer minutes as their foul count increases (Chu and Swartz, 2020), the sample size of shots taken in the presence of a defender with a high foul count versus a low foul count is unbalanced. Multilevel models and adaptive regularizing priors (McElreath, 2020, pp. 408) are used to mitigate over fitting such that comparisons can be made across foul counts.

# Literature Review

Bayesian models have been used in a number of previous studies to model the both offensive and defensive performance. Williams et al. (2024) used a Bayesian hierarchical modeling framework to model offensive performance of teams and players based on their shooting propensities and abilities. They used discretized shot loactions for the top 100 shot-takers from the 2008/2009–2020/2021 seasons to create an expected points above average (EPAA) metric. Our project differs in our use of hierarchical modeling to infer defensive impact, rahter than contributing to the existing research on offense.

Franks et al. (2015) was one of the pioneering papers in using Bayesian hierarchical models to study player tracking data by modeling the relationship between a defensive assignment and the spacial-temporal shotmaking tendencies of various offensive players, which is to say how well they shoot the ball given the location, time, and defender of a shot attempt. Our project differs in that we use a Bayesian hierarchical model to infer defensive impact without using raw player tracking data, rather than modeling the relationship between a known defensive assignment and shotmaking tendencies.

Finally, Chu and Swartz (2020) use Bayesian methods to model the time-to-event of a foul as a gamma distributed function of a player, their accumulated number of fouls, and their position. Their goal was to develop a decision-making framework for coaches to use in determining when to sit players at risk of fouling out (usually when a player reaches 6 fouls) of a game. Although similar in spirit, our project doesn't model the time-to-event of a foul nor attempt to help coaches make decisions about sitting their players, but instead focuses on the effect of fouls while players are on the court.

# Modeling Strategy

This section details the steps taken to develop a novel model for inferring player defensive skill based on the Bayesian workflow proposed by Gelman and Hill (BDA3, 2006) and others. Naming conventions and statistical notation are in line with those used by McElreath.

It includes: - a narrative model of the system of inquiry (Betancourt, TK), - a heuristic causal model motivated by the narrative, - an agent-based generative model used to simulate sythetic observations of shots within a basketball game, - a mathematical model, and - a computational model compatible with the Hamiltonian Monte-Carlo sample implemented in the Stan probabilistic programming language (Carpenter, 2017).

## Narrative Model

\[This section is a work in progress. I plan on taking out the pseudo math, which is currently in place to help me think through the causal model\]

We assume that basketball is essentially a game of occupying physical space. We assume that at any given time T, each location L on a court of dimension TKxTK has some value V to an offensive player OP, who makes decisions so as to maximize V. As in chess, defenders are advantaged by anticipating the spatial preferences (V \| L) of their opponents and at time T-1 occupying the location L where V is maximized.

Any given location L can be occupied by no more than one player (offensive or defensive) at any given time T.

Essentially, at any given time, offensive players prefer to occupy whatever location on the court maximizes their probability of their team scoring the basket. Defensive players are able to reduce the value of any location in one of two ways:

1.  By using his defensive skill to preemptively occupy the highest value max(V \| T) location, thereby reducing the total offensive value of any available location.

2.  Having been unable to preemptively occupy the highest value max(V \| T), by using defensive pressure to initiate contact at max(V \| T), thereby lowering its value directly. This could look functionally something like V \| T \~ f(max(V \| T) + (dp \| T))

This assumes that defensive players' ability to affect shotmaking probability for any given L is some additive function of his **defensive pressure** and his **defensive skill**.

We can then assume that defenders with more defensive skill require less defensive pressure to affect shotmaking probability for any max(V \| T).

We also assume that defenders who use too much defensive pressure to affect the value of max(V \| T) are assessed a foul, which accumulate at some rate over the course of a given game. E.g. foul_assessed \~ f(max(V \| T) - V \| T \>)

We also assume that as a given player's fouls increase, his ability to express physical skill drops, and so a greater proportion of his ability to affect shotmaking is comprised of his defensive skill.

The difference between a given defensive player's affect on the probability that a shot is made when the shot is taken while that defensive player has some number of fouls versus some other number of fouls can be interpreted as the total amount of his shot-affecting ability is owed to his defensive skill versus his defensive pressure.

For an illustration of the narrative model in action we can think about the offensive charge, a case in which a defender anticipates the movements of a driving offensive player to such an accurate degree that he's able to beat that offender to the location of his shot, resulting in contact initiated entirely by the offender that is ultimately called a turnover in favor of the defense. Taking a charge as it's called (perhaps we should think of it as causing a charge) requires no physical pressure on the part of the defender, who instead uses his knowledge of the game and IQ (plausible latent traits of a good defender) to gain the advantage on a possession.

Or, we can consider the defensive move known as "pulling the chair" in the post. In this scenario, a defender anticipates that an offensive player (facing away from the basket) is about to initiate contact to establish position. At the last moment, the defender steps away, causing the offensive player to lose balance. Like drawing a charge, this move relies primarily on defensive skill (anticipation and timing) rather than physical pressure. If pulling the chair effectively reduces shotmaking probability by disrupting the offensive player's balance, we would expect to see minimal correlation between foul accumulation and defensive effectiveness on post-up shots for players who excel at this technique.

## Causal model

The complete causal model assumed in this paper can be visualized with the following Directed Acyclic Graph (DAG):

{{< include ../03_causal-model/causal-model_final.qmd >}}

This causal diagram includes measurable variables and unmeasurable, latent player traits which we may infer using an assumed narrative

# Generative Model

The [complete generative model](https://github.com/MichaelPaulLight/nba_foul-analysis/blob/main/00_R/simulate_shots.R) used for prior elicitation and model validation via simulation based calibration can be found in the associated materials.

```{r}
source(here::here("00_R", "simulate_shots.R"))

small_sim <- simulate_shots(n_observations = 1000, n_defenders = 50, n_teams = 10) |> head(5)

small_sim |> gt()
```

# Statistical Models

The statistical model can be written as the following:

# Prior Selection and Predictive Checks

## Two Prior Models for each Model of Foul Effect on Shot-taking

### Less Constrained

### More Constrained

# Model Validation

## Simulation Based Calibration

# Observed Sample Selection

## Observed Sample of Shots (Play-by-Play)

# Model Fitting and Diagnostics

## Fitting Models to Observed Samples

## HMC Diagnostics

# Posterior Analysis

## Posterior Retrodictive Checks

## Posterior Residual Checks

# Counter-factual Analysis

# Conclusion

# Appendices

## Appendix A.

### References

Add:

-   BDA 3 (Gelman and Hill)
-   Statistical rethinking (McElreath)
-   Stan (Carpenter, 2017)
-   Betancourt narrative model

Chu, D., & Swartz, T. (2020). Foul accumulation in the NBA. Journal of Quantitative Analysis in Sports, 16(4), 301-309. https://doi.org/10.1515/jqas-2019-0119

Franks, A., Miller, A., Bornn, L., Goldsberry, K., Characterizing the spatial structure of defensive skill in professional basketball. (2015) Ann. Appl. Stat. 9 (1) 94 - 121. https://doi.org/10.1214/14-AOAS799

Simmons, B. (2009). The book of basketball: the NBA according to the sports guy. Ballantine/ESPN Books.

Williams, B., Schliep, E. M., Fosdick, B., & Elmore, R. (2024). Expected points above average: A novel NBA player metric based on Bayesian hierarchical modeling. arXiv preprint arXiv:2405.10453. https://arxiv.org/abs/2405.10453
