```{r}
library(hoopR)
library(tidyverse)
library(nanoparquet)
library(janitor)
library(zoo)
library(here)
library(duckplyr)
```

This notebook investigates foul count inflation in the 2024-25 NBA dataset as a possible data quality issue.

Ultimately, it was determined that NBA play-by-play data sometimes contains inconsistencies in personal foul counts. Data cleaning steps to handle erroneuos counts are conducted in clean_pbp.qmd, and summarized in here("02_data", "02-2_data_rclean", "overall_foul_correction_summary.parquet")

```{r}
pbp_data <- read_parquet(here("02_data", "02-1_data_raw", "250409_pbp_gt.parquet"))

pbp_data <- pbp_data %>%
  mutate(fouling_player = case_when(
    str_detect(description, regex("foul:", ignore_case = T)) ~ player1, 
    .default = NA
  )) %>%
  mutate(fouled_player = case_when(
    str_detect(description, regex("foul:", ignore_case = T)) ~ player3, 
    .default = NA
  )) %>%
  mutate(personal_fouls_after_event = case_when(
    !is.na(fouling_player) ~ str_extract(description, regex("\\([^()]*\\)(?![^()]*\\))"))
  )) %>%
  mutate(personal_fouls_after_event = str_extract(personal_fouls_after_event, regex("\\d+")))


# Check for foul count inconsistencies
foul_sequence_check <- pbp_data %>%
  filter(!is.na(fouling_player)) %>%
  group_by(game_id, fouling_player) %>%
  arrange(secs_passed_game) %>%
  mutate(
    foul_count = as.numeric(personal_fouls_after_event),
    prev_foul_count = lag(foul_count),
    foul_count_diff = foul_count - coalesce(prev_foul_count, 0),
    is_inconsistent = case_when(
      is.na(foul_count) ~ FALSE,
      foul_count < 0 ~ TRUE,
      foul_count > 6 ~ TRUE,
      !is.na(prev_foul_count) & foul_count_diff != 1 ~ TRUE,
      .default = FALSE
    )
  ) %>%
  filter(is_inconsistent) %>%
  select(game_id, fouling_player, secs_passed_game, description, foul_count, prev_foul_count, foul_count_diff)

# View the inconsistencies
foul_sequence_check

# Count the number of games with inconsistencies
foul_sequence_check %>%
  distinct(game_id) %>%
  count()

# Count the number of players with inconsistencies
foul_sequence_check %>%
  distinct(game_id, fouling_player) %>%
  count()
```

```{r}
nonagg_data <- read_parquet(here("02_data", "02-2_data_clean", "shot_level_defender_matchup_data_all.parquet"))

investigation <- read_parquet(here("02_data", "02-2_data_clean", "overall_foul_correction_summary.parquet"))

foul_jumps <- nonagg_data %>%
  ungroup() |> 
  # Group by game and defender
  group_by(game_id, primary_defender) %>%  # Removed play_number from grouping
  # Sort by play number within each group
  arrange(game_id, primary_defender, play_number) %>%
  # Calculate the difference in foul count from previous play
  mutate(
    prev_foul_count = lag(defender_foul_count, order_by = play_number)) |> 
  mutate(prev_foul_count = replace_na(prev_foul_count, 0),
    foul_count_diff = defender_foul_count - prev_foul_count
  ) %>%
  # Filter for cases where the foul count increases by more than 1
  filter(!is.na(foul_count_diff) & foul_count_diff > 1)

foul_jumps |> ungroup() |> select(game_id) |> distinct()

  # Select relevant columns
  select(
    game_id, 
    primary_defender, 
    play_number, 
    prev_play_number = lag(play_number),
    defender_foul_count, 
    prev_foul_count, 
    foul_count_diff
  )

# Summary of findings
jump_summary <- foul_jumps %>%
  summarise(
    n_jumps = n(),
    n_games = n_distinct(game_id),
    n_players = n_distinct(primary_defender),
    max_jump = max(foul_count_diff, na.rm = TRUE)
  )

# Print results
print("Summary of foul count jumps greater than 1:")
print(jump_summary)

```

