```{r}
library(hoopR)
library(tidyverse)
library(nanoparquet)
library(janitor)
library(zoo)
library(here)
library(duckplyr)
```

This notbook takes our non-aggregated shot-level data and aggregates it for use with aggregated binomial models.

```{r}
nonagg_data <- read_parquet(here("02_data", "02-2_data_clean", "shot_level_defender_matchup_data_all.parquet"))
```

```{r}
nonagg_data |> filter(is.na(defender_position)) |> select(primary_defender) |> distinct()

```


Because of the way roster data is reported by NBA teams, there are some players for whom we lack position information. For the purpose of this analysis, we'll drop these cases.

We'll also split the shot_action variable into less-specific groupings: shot_movement and shot_type. 

```{r}
aggregated_data <- nonagg_data |> 
  group_by(game_id, period, slug_team_off, slug_team_def, primary_defender, shot_action, shot_zone, shot_area, shot_distance_ft, shot_type, defender_position, defender_foul_count) |> 
  summarize(shots_defended_made = sum(shot_made),
         shots_defended_total = n()) |> 
  rename(defensive_team = slug_team_def,
         offensive_team = slug_team_off) |> 
  filter(!is.na(defender_position))

aggregated_data <- aggregated_data |> 
  mutate(
    # Shot movement - how the player is moving
    shot_movement = case_when(
      str_detect(shot_action, regex("driving", ignore_case = TRUE)) ~ "driving",
      str_detect(shot_action, regex("running", ignore_case = TRUE)) ~ "running", 
      str_detect(shot_action, regex("step back", ignore_case = TRUE)) ~ "step back",
      str_detect(shot_action, regex("pullup|pull-up", ignore_case = TRUE)) ~ "pullup",
      str_detect(shot_action, regex("cutting", ignore_case = TRUE)) ~ "cutting",
      str_detect(shot_action, regex("floating", ignore_case = TRUE)) ~ "floating",
      TRUE ~ "stationary"
    ),
    
    # Shot type - fundamental shot category
    shot_type_basic = case_when(
      str_detect(shot_action, regex("layup", ignore_case = TRUE)) ~ "layup",
      str_detect(shot_action, regex("dunk", ignore_case = TRUE)) ~ "dunk",
      str_detect(shot_action, regex("jump shot|jumpshot", ignore_case = TRUE)) ~ "jump shot",
      str_detect(shot_action, regex("hook", ignore_case = TRUE)) ~ "hook shot",
      str_detect(shot_action, regex("finger roll", ignore_case = TRUE)) ~ "finger roll",
      str_detect(shot_action, regex("tip", ignore_case = TRUE)) ~ "tip shot",
      TRUE ~ "other"
    ),
    
    # Shot technique - style or special technique used
    shot_technique = case_when(
      str_detect(shot_action, regex("fadeaway", ignore_case = TRUE)) ~ "fadeaway",
      str_detect(shot_action, regex("finger roll", ignore_case = TRUE)) ~ "finger roll",
      str_detect(shot_action, regex("hook", ignore_case = TRUE)) ~ "hook",
      str_detect(shot_action, regex("turnaround", ignore_case = TRUE)) ~ "turnaround",
      str_detect(shot_action, regex("floating", ignore_case = TRUE)) ~ "floating",
      TRUE ~ "standard"
    ),
    
    # Shot angle - relation to basket
    shot_angle = case_when(
      str_detect(shot_action, regex("reverse", ignore_case = TRUE)) ~ "reverse",
      str_detect(shot_action, regex("bank", ignore_case = TRUE)) ~ "bank",
      TRUE ~ "standard"
    ),
    
    # Shot situation - contextual situation
    shot_situation = case_when(
      str_detect(shot_action, regex("putback", ignore_case = TRUE)) ~ "putback",
      str_detect(shot_action, regex("alley oop", ignore_case = TRUE)) ~ "alley oop",
      str_detect(shot_action, regex("tip", ignore_case = TRUE)) ~ "tip",
      str_detect(shot_action, regex("cutting", ignore_case = TRUE)) ~ "cutting",
      TRUE ~ "standard"
    )
  )

# Function to create and save samples
create_and_save_samples <- function(sample_size) {
  set.seed(123)
  
  # Create strata
  data_sample <- aggregated_data |> 
    # First sample games
    group_by(game_id) |> 
    sample_frac(size = 1)  |>
    ungroup() |> 
    # Then sample within games
    group_by(period, defensive_team) |> 
    sample_frac(size = sample_size/nrow(aggregated_data), replace = FALSE) |>
    mutate(defender_foul_count = as.numeric(defender_foul_count)) |> 
    ungroup()
  
  # Create scaled version
  data_sample_scaled <- data_sample |> 
    mutate(defender_foul_count = scale(defender_foul_count)[,1])
  
  # Create ordered factor version
  data_sample_factor <- data_sample |> 
    mutate(defender_foul_count = factor(defender_foul_count, 
                                   levels = 0:5,
                                   ordered = TRUE))
  
  # Save the samples
  write_parquet(data_sample, 
                here("02_data", "02-3_data_modeling", 
                     paste0("sample_", sample_size, ".parquet")))
  write_parquet(data_sample_scaled, 
                here("02_data", "02-3_data_modeling", 
                     paste0("sample_", sample_size, "_scaled.parquet")))
  write_parquet(data_sample_factor, 
                here("02_data", "02-3_data_modeling", 
                     paste0("sample_", sample_size, "_factor.parquet")))
  
  return(nrow(data_sample))
}

# Create samples of different sizes
sample_sizes <- c(5000, 10000, 30000, 60000, 100000, 200000, nrow(aggregated_data))

# Create and save all samples
sample_counts <- map(sample_sizes, create_and_save_samples)

# Print sample sizes for verification
print("Sample sizes created:")
print(data.frame(
  target_size = sample_sizes,
  actual_size = unlist(sample_counts)
))
```
